---
title: "GDP and life expectancy in 2007"
author: "Wei Huang"
date: "2026-09-22"
---
Data: [Gapminder](https://www.gapminder.org/data/), via the
  [gapminder R package](https://github.com/jennybc/gapminder). Licensed CC BY 4.0.

The gapminder dataset contains life expectancy, population, and GDP per capita data for 142 countries from 1952 to 2007. In this post, I explore the correlation between GDP per capita and life expectancy, using data from 2007, the most recent year in the dataset.

```{r}
#| label: setup
#| message: false
#| warning: false
  library(gapminder)
  library(dplyr)
  library(ggplot2)
```

## Step 1: Filter the data
First, I filter the data to keep only the records from 2007.
```{r}
#| label: filter-2007
gap07 <- gapminder |>
    filter(year == 2007)
gap07
```

## Step 2: Compare median GDP and life expectancy by continent
Next, I compare the median GDP and life expectancy by continent. Oceania has the highest median GDP and life expectancy, while Africa has the lowest on both.
```{r}
#| label: continent-comparison
comparison <- gap07 |>
    group_by(continent) |>
    summarise(
        countries = n(),
        median_gdp = round(median(gdpPercap), 2),
        median_lifeExp = round(median(lifeExp), 2)
    )
comparison
```

## Step 3: Correlation
Then I calculate the correlation between these two variables. The result is 0.68, a fairly strong positive relationship.
```{r}
#| label: correlations
corr <- cor(gap07$gdpPercap, gap07$lifeExp)
corr
```

## Step 4: Visualization
Finally, I create a scatter plot that shows this positive relationship, with the points coloured by continent to tell them apart. Note that since we have not yet learned about visualization, this snippet is a direct copy from Claude.
```{r}
#| label: fig-lifeexp-gdp
#| fig-cap: "Life expectancy vs. GDP per capita by country, 2007."
   ggplot(gap07, aes(x = gdpPercap, y = lifeExp, colour = continent)) +
    geom_point() +
    labs(x = "GDP per capita (US$)", y = "Life expectancy (years)")
```

The plot shows an interesting pattern. Although the overall relationship is positive, life expectancy rises rapidly among low-income countries, then levels off and remains stable among middle- and high-income countries. Meanwhile, countries with similarly low incomes can have very different life expectancies (e.g., countries in Africa versus Asia). This suggests that other factors besides GDP per capita have a strong influence on life expectancy.